---
title: "Machine Learning Project"
output: html_document
date: "2025-03-20"
---
I want to emphasize all functions have been created by ChatGpt. I used it to help debug and also explain each line of code so someone can easily read through it!

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Data and packages

```{r}
# Disable scientific notation
options(scipen = 999)

# Set number of digits to display (3 decimal places)
options(digits = 3)
data <-read.table("C:/Users/ashvi/OneDrive/Desktop/ML Business Forecasting/data_tr.txt", head=T)[,-1]

library("splines")
library("tidyverse")
library("glmnet")
library("leaps")
library("boot")
library("MASS")
library("SemiPar")
library("ISLR")
library("mgcv")
```

## Understanding Data Format

Total Wealth can go negative, people can go in debt. 
Income can be
negative(garnished wages perhaps) 
1st Quarter people home values are less then their mortgage. 
Dataset only goes up to those who are 65. 
Two earners and being elegible for IRA is only a small percentage of people

```{r}
dim(data)
head(data)
names(data)
summary(data)
```

##Visualize data relationships 
IRA-positive 
e401-positive 
nifa-positive
inc-slightly positive 
hmort-positive but also weirdly flat 
hval-positive
hequity-positive
educ-positive but small magnitude
male-positive 
twoearn-1 earner seems to make slightly more education
dummies-seems a lot more stronger, will examine relationship to create
additional variable for those who did more education then college
Age-slightly positive family size-non linear relationship, peaks around
2 then diminishes married-significant variation compared to non married
individuals

```{r}
# Define the directory to save plots
save_dir <- "C:/Users/ashvi/Downloads/"

# Loop to plot and save each relationship of data
for (i in 2:18) {

  
  # Create the plot
  plot(data[, i], data[, 1], 
       xlab = names(data)[i], 
       ylab = names(data)[1], 
       main = paste("TW vs", names(data)[i]))
}
  
```
Here I will use 2 outlier removal methods. First will use  transformation to reduce extreme values in wealth and income. Then I will be using IQR and removing variables that fall outside the typical range(25%-75%).  Finally I will will be using z scores to standardize the data. Secound method will be dropping observations greater then 1,000,000 from total wealth. 
```{r}
#Visualize numeric data
hist(data$tw)
hist(data$inc)

# Function to remove outliers using IQR method
remove_outliers_iqr <- function(x) {
  Q1 <- quantile(x, 0.25)
  Q3 <- quantile(x, 0.75)
  IQR <- Q3 - Q1
  lower_bound <- Q1 -  16* IQR
  upper_bound <- Q3 + 16* IQR
  return(x >= lower_bound & x <= upper_bound)
}

# Apply outlier removal for `tw`, and `inc`
outlier_mask <- sapply(data[, c( "tw", "inc")], function(x) remove_outliers_iqr(x))

# Track the number of rows dropped due to outliers
rows_before_outliers <- nrow(data)
data_clean1 <- data[apply(outlier_mask, 1, all), ]
rows_after_outliers <- nrow(data_clean1)
dropped_rows_outliers <- rows_before_outliers - rows_after_outliers
cat("Rows dropped due to outliers in `ira`, `tw`, and `inc`: ", dropped_rows_outliers, "\n")

# Drop rows where `tw` is greater than 1,000,000 from `data`
rows_before_tw_drop <- nrow(data)  # Count rows before filtering
data_clean2 <- data[data$tw <= 1000000, ]  # Correct filtering
rows_after_tw_drop <- nrow(data_clean2)  # Count rows after filtering
dropped_rows_tw <- rows_before_tw_drop - rows_after_tw_drop  # Compute number of dropped rows

# Print the number of dropped rows
cat("Rows dropped due to `tw > 1,000,000`: ", dropped_rows_tw, "\n")

hist(data_clean1$tw)
hist(data_clean1$inc)

```


##Baseline Model
hequity=hval-hmort
education variables dummies, watch out for collinearity
Here is the baseline model trained on the three datasets and tested on the base dataset. The least squares automatically dropped perfectly collinear variables(hequity and college dummy)

```{r}
# Define number of folds for K-Fold Cross-Validation
k <- 10
n <- nrow(data)

# Set seed for reproducibility
set.seed(7141)

# Generate k-fold indices
id <- sample(rep(1:k, length = n))

# Initialize vectors to store MSPE results for each model
MSPE_0 <- MSPE_1 <- MSPE_2 <- vector(length = k)

# Perform 10-fold cross-validation
for (f in 1:k) {
  train <- (id != f)  # Training set indices
  test <- (id == f)   # Test set indices

  ## Train models on different datasets
  model0 <- lm(tw ~ ., data = data[train,])         # Model trained on original  without log transformations
  model1 <- lm(tw ~ ., data = data_clean1[train,])   # Model trained on cleaned data (log + IQR)
  model2 <- lm(tw ~ ., data = data_clean2[train,])   # Model trained on wealth-capped data

  ## Make predictions on the original test set (data[test,])
  pr_0 <- predict(model0, newdata = data[test,])
  pr_1 <- predict(model1, newdata = data[test,])
  pr_2 <- predict(model2, newdata = data[test,])

  ## Compute MSPE (Mean Squared Prediction Error) on `data`
  MSPE_0[f] <- mean((pr_0 - data$tw[test])^2)
  MSPE_1[f] <- mean((pr_1 - data$tw[test])^2)
  MSPE_2[f] <- mean((pr_2 - data$tw[test])^2)
}

# Compute the average MSPE for each model
mean_MSPE_0 <- mean(MSPE_0)
mean_MSPE_1 <- mean(MSPE_1)
mean_MSPE_2 <- mean(MSPE_2)

# Print results
cat("Baseline Model MSPE (Trained on Original Data):", mean_MSPE_0, "\n")
cat("Model 1 MSPE (Trained on IQR Cleaned Data):", mean_MSPE_1, "\n")
cat("Model 2 MSPE (Trained on Wealth Cap Removed Data):", mean_MSPE_2, "\n")

```
##Ridge/Lasso/Stepwise

Stepwise Regression

```{r}
# Define the full and null model
full <- lm(tw ~ ., data=data)
null <- lm(tw ~ 1, data=data)

# forward stepwise - AIC
a <- stepAIC(null, scope=list(lower=null, upper=full), trace = F, direction='forward')

# backward stepwise - AIC
b <- stepAIC(full, scope=list(lower=null, upper=full), trace = F, direction='backward')

# Look at the coefficients to know which variables were selected
coef(a)
coef(b)

# Predict forward and backward
pr.stepwise_backward <- predict(b, newdata=data)
pr.stepwise_forward <- predict(a, newdata=data)
```
Ridge Regression

```{r}
#choosing sequence of lambda
lambdas.rr <- exp(seq(-20, 20, length = 200))
#seperating to training and testing sets
y <- data$tw
X <- as.matrix(data[,-1])

#set seed
set.seed(67134)

#Run ridge regression
ridge_cv <- cv.glmnet(x = X, y = y, lambda = lambdas.rr,alpha = 0)
# value of lambda that gives minimum mean cross-validated error
ridge_cv$lambda.min

# Use lamda.min to get the final model:
ridge <- glmnet(x = X, y = y, lambda = ridge_cv$lambda.min,alpha = 0)

```
Lasso Regression
```{r}
#choosing sequence of lambda
lambdas.LASSO <- exp(seq(-20, 20, length = 200))
set.seed(123)
LASSO_cv <- cv.glmnet(x = X, y = y, lambda = lambdas.LASSO,alpha = 1)
# value of lambda that gives minimum mean cross-validated error
LASSO_cv$lambda.min
#use best lamnda
LASSO <- glmnet(x = X, y = y, lambda = LASSO_cv$lambda.min,alpha = 1)
LASSO$beta
plot(LASSO_cv)
```


K-Fold Cross Validation Lasso v Ridge v Stepwise

```{r}
# 10-fold cross-validation setup
k <- 10
n <- nrow(data)
set.seed(67134)
id <- sample(rep(1:k, length = n))

# Initialize vectors to store MSPE for each method
MSPE_stepwise_forward <- MSPE_stepwise_backward <- MSPE_ridge <- MSPE_lasso <- vector(length = k)

# Perform cross-validation
for (f in 1:k) {
  train <- (id != f)  # Training set
  test <- (id == f)   # Test set
  
  # Prepare training and test data
  X_train <- as.matrix(data[train, -1])  # Excluding the target variable `tw`
  y_train <- data$tw[train]
  X_test <- as.matrix(data[test, -1])
  y_test <- data$tw[test]
  
  # 1. Forward Stepwise
  full_model <- lm(tw ~ ., data = data[train,])
  null_model <- lm(tw ~ 1, data = data[train,])
  forward_model <- stepAIC(null_model, scope = list(lower = null_model, upper = full_model), direction = "forward", trace = FALSE)
  pr_forward <- predict(forward_model, newdata = data[test,])
  MSPE_stepwise_forward[f] <- mean((pr_forward - y_test)^2)
  
  # 2. Backward Stepwise
  backward_model <- stepAIC(full_model, scope = list(lower = null_model, upper = full_model), direction = "backward", trace = FALSE)
  pr_backward <- predict(backward_model, newdata = data[test,])
  MSPE_stepwise_backward[f] <- mean((pr_backward - y_test)^2)
  
  # 3. Ridge Regression
  ridge_cv <- cv.glmnet(x = X_train, y = y_train, alpha = 0)  # Ridge regression
  ridge_model <- glmnet(x = X_train, y = y_train, lambda = ridge_cv$lambda.min, alpha = 0)
  pr_ridge <- predict(ridge_model, newx = X_test)
  MSPE_ridge[f] <- mean((pr_ridge - y_test)^2)
  
  # 4. Lasso Regression
  lasso_cv <- cv.glmnet(x = X_train, y = y_train, alpha = 1)  # Lasso regression
  lasso_model <- glmnet(x = X_train, y = y_train, lambda = lasso_cv$lambda.min, alpha = 1)
  pr_lasso <- predict(lasso_model, newx = X_test)
  MSPE_lasso[f] <- mean((pr_lasso - y_test)^2)
}

# Calculate the average MSPE for each model
avg_MSPE_stepwise_forward <- mean(MSPE_stepwise_forward)
avg_MSPE_stepwise_backward <- mean(MSPE_stepwise_backward)
avg_MSPE_ridge <- mean(MSPE_ridge)
avg_MSPE_lasso <- mean(MSPE_lasso)

# Print the results
cat("Stepwise Forward MSPE:", avg_MSPE_stepwise_forward, "\n")
cat("Stepwise Backward MSPE:", avg_MSPE_stepwise_backward, "\n")
cat("Ridge Regression MSPE:", avg_MSPE_ridge, "\n")
cat("Lasso Regression MSPE:", avg_MSPE_lasso, "\n")
```
Now I will repeat this by training on the new clean datasets and testing on the "original" training dataset

```{r}
## T Train models on data_clean1
set.seed(67134)
# Stepwise Forward Model
full_model_clean1 <- lm(tw ~ ., data = data_clean1)
null_model_clean1 <- lm(tw ~ 1, data = data_clean1)
forward_model_clean1 <- stepAIC(null_model_clean1, scope = list(lower = null_model_clean1, upper = full_model_clean1), direction = "forward", trace = FALSE)

# Stepwise Backward Model
backward_model_clean1 <- stepAIC(full_model_clean1, scope = list(lower = null_model_clean1, upper = full_model_clean1), direction = "backward", trace = FALSE)

# Ridge Regression
X_clean1 <- as.matrix(data_clean1[,-1])
y_clean1 <- data_clean1$tw
ridge_cv_clean1 <- cv.glmnet(x = X_clean1, y = y_clean1, alpha = 0)
ridge_model_clean1 <- glmnet(x = X_clean1, y = y_clean1, lambda = ridge_cv_clean1$lambda.min, alpha = 0)

# Lasso Regression
lasso_cv_clean1 <- cv.glmnet(x = X_clean1, y = y_clean1, alpha = 1)
lasso_model_clean1 <- glmnet(x = X_clean1, y = y_clean1, lambda = lasso_cv_clean1$lambda.min, alpha = 1)

# Test on the whole dataset data
# Prepare the test data (original data, not cleaned)
X_data <- as.matrix(data[,-1])  # Use the full dataset
y_data <- data$tw

# Predict using the forward stepwise model (trained on data_clean1)
pr_forward_clean1 <- predict(forward_model_clean1, newdata = data)
MSPE_stepwise_forward_clean1 <- mean((pr_forward_clean1 - y_data)^2)

# Predict using the backward stepwise model (trained on data_clean1)
pr_backward_clean1 <- predict(backward_model_clean1, newdata = data)
MSPE_stepwise_backward_clean1 <- mean((pr_backward_clean1 - y_data)^2)

# Predict using Ridge model (trained on data_clean1)
pr_ridge_clean1 <- predict(ridge_model_clean1, newx = X_data)
MSPE_ridge_clean1 <- mean((pr_ridge_clean1 - y_data)^2)

# Predict using Lasso model (trained on data_clean1)
pr_lasso_clean1 <- predict(lasso_model_clean1, newx = X_data)
MSPE_lasso_clean1 <- mean((pr_lasso_clean1 - y_data)^2)

# Print MSPE for the models trained on data_clean1 and tested on full data
cat("Forward Stepwise MSPE (trained on data_clean1, tested on data):", MSPE_stepwise_forward_clean1, "\n")
cat("Backward Stepwise MSPE (trained on data_clean1, tested on data):", MSPE_stepwise_backward_clean1, "\n")
cat("Ridge Regression MSPE (trained on data_clean1, tested on data):", MSPE_ridge_clean1, "\n")
cat("Lasso Regression MSPE (trained on data_clean1, tested on data):", MSPE_lasso_clean1, "\n")

# Now, repeat the process for data_clean2
# Train models on data_clean2 (similar to the process for data_clean1)
# Stepwise Forward Model
full_model_clean2 <- lm(tw ~ ., data = data_clean2)
null_model_clean2 <- lm(tw ~ 1, data = data_clean2)
forward_model_clean2 <- stepAIC(null_model_clean2, scope = list(lower = null_model_clean2, upper = full_model_clean2), direction = "forward", trace = FALSE)

# Stepwise Backward Model
backward_model_clean2 <- stepAIC(full_model_clean2, scope = list(lower = null_model_clean2, upper = full_model_clean2), direction = "backward", trace = FALSE)

# Ridge Regression
X_clean2 <- as.matrix(data_clean2[,-1])
y_clean2 <- data_clean2$tw
ridge_cv_clean2 <- cv.glmnet(x = X_clean2, y = y_clean2, alpha = 0)
ridge_model_clean2 <- glmnet(x = X_clean2, y = y_clean2, lambda = ridge_cv_clean2$lambda.min, alpha = 0)

# Lasso Regression
lasso_cv_clean2 <- cv.glmnet(x = X_clean2, y = y_clean2, alpha = 1)
lasso_model_clean2 <- glmnet(x = X_clean2, y = y_clean2, lambda = lasso_cv_clean2$lambda.min, alpha = 1)

# Test on the whole dataset data
# Predict using the forward stepwise model (trained on data_clean2)
pr_forward_clean2 <- predict(forward_model_clean2, newdata = data)
MSPE_stepwise_forward_clean2 <- mean((pr_forward_clean2 - y_data)^2)

# Predict using the backward stepwise model (trained on data_clean2)
pr_backward_clean2 <- predict(backward_model_clean2, newdata = data)
MSPE_stepwise_backward_clean2 <- mean((pr_backward_clean2 - y_data)^2)

# Predict using Ridge model (trained on data_clean2)
pr_ridge_clean2 <- predict(ridge_model_clean2, newx = X_data)
MSPE_ridge_clean2 <- mean((pr_ridge_clean2 - y_data)^2)

# Predict using Lasso model (trained on data_clean2)
pr_lasso_clean2 <- predict(lasso_model_clean2, newx = X_data)
MSPE_lasso_clean2 <- mean((pr_lasso_clean2 - y_data)^2)

# Print MSPE for the models trained on data_clean2 and tested on full data
cat("\nForward Stepwise MSPE (trained on data_clean2, tested on data):", MSPE_stepwise_forward_clean2, "\n")
cat("Backward Stepwise MSPE (trained on data_clean2, tested on data):", MSPE_stepwise_backward_clean2, "\n")
cat("Ridge Regression MSPE (trained on data_clean2, tested on data):", MSPE_ridge_clean2, "\n")
cat("Lasso Regression MSPE (trained on data_clean2, tested on data):", MSPE_lasso_clean2, "\n")
```
Creating every possible interaction term
```{r}
# Remove 'tw' from the dataset since it's the target variable
data_no_tw <- data[, !colnames(data) %in% "tw"]

# Get the original feature names
original_features <- names(data_no_tw)

# Generate all possible pairwise interaction terms
interaction_terms <- combn(original_features, 2, FUN = function(x) paste(x[1], x[2], sep=":"), simplify = TRUE)

# Create the formula for the linear model including both original features and interaction terms
formula_terms <- paste("tw ~", paste(c(original_features, interaction_terms), collapse=" + "))

# Print the formula
print(formula_terms)
```
Using feature selection methods to chose the best combination of variables
```{r}
# Fit full model with interaction terms
full_model <- lm(as.formula(formula_terms), data = data)

# Null model (intercept only)
null_model <- lm(tw ~ 1, data = data)

# Perform stepwise selection using both forward and backward directions based on AIC
stepwise_model <- stepAIC(null_model, scope = list(lower = null_model, upper = full_model), direction = "both", trace = FALSE)

# View selected features from the stepwise model
cat("Selected features in Stepwise Model:\n")
summary(stepwise_model)
```

```{r}
# Create the model matrix (predictors)
X_interactions <- model.matrix(as.formula(formula_terms), data = data)[,-1]  # Remove the intercept column
y <- data$tw  # Target variable

# Perform Lasso regression with cross-validation to select the optimal lambda
lasso_cv <- cv.glmnet(X_interactions, y, alpha = 1)  # alpha=1 for Lasso regression

# Best lambda selected by cross-validation
cat("Best lambda from Lasso CV:", lasso_cv$lambda.min, "\n")

# Fit Lasso model using the best lambda
lasso_model <- glmnet(X_interactions, y, lambda = lasso_cv$lambda.min, alpha = 1)

# View the coefficients of the Lasso model
cat("\nLasso Regression Coefficients:\n")
coef(lasso_model)
summary(lasso_model)
```
Select features from interaction dataset
```{r}
# Stepwise Model (already fitted, e.g., stepwise_model)
stepwise_coefficients <- coef(stepwise_model)
selected_features_stepwise <- names(stepwise_coefficients)[stepwise_coefficients != 0]
selected_features_stepwise <- selected_features_stepwise[selected_features_stepwise != "(Intercept)"]

cat("Selected features from Stepwise model:\n")
print(selected_features_stepwise)

# Lasso Model (assuming lasso_model is already fitted)
lasso_coefficients <- coef(lasso_model)
lasso_coefficients_matrix <- as.matrix(lasso_coefficients)
selected_features_lasso <- rownames(lasso_coefficients_matrix)[lasso_coefficients_matrix != 0]
selected_features_lasso <- selected_features_lasso[selected_features_lasso != "(Intercept)"]

cat("Selected features from Lasso model:\n")
print(selected_features_lasso)

```

Use cross validation to see which model feature selection was better with ridge regression
```{r}
# Function to calculate MSPE (Mean Squared Prediction Error)
calculate_mspe <- function(model, X_test, y_test) {
  predictions <- predict(model, newx = X_test)
  return(mean((predictions - y_test)^2))
}

# Set the number of folds for cross-validation (10-fold)
k <- 10
n <- nrow(data_clean1)  
set.seed(67134)
id <- sample(rep(1:k, length = n))

# Prepare Stepwise-selected features in `data_clean1`
X_clean1_stepwise <- model.matrix(as.formula(paste("tw ~", paste(selected_features_stepwise, collapse = " + "))), data = data_clean1)[, -1]
# Prepare Lasso-selected features in `data_clean1`
X_clean1_lasso <- model.matrix(as.formula(paste("tw ~", paste(selected_features_lasso, collapse = " + "))), data = data_clean1)[, -1]
y_clean1 <- data_clean1$tw

# Prepare test set from `data`
X_test_stepwise <- model.matrix(as.formula(paste("tw ~", paste(selected_features_stepwise, collapse = " + "))), data = data)[, -1]
X_test_lasso <- model.matrix(as.formula(paste("tw ~", paste(selected_features_lasso, collapse = " + "))), data = data)[, -1]
y_test <- data$tw

# Initialize vectors to store MSPE for each fold
MSPE_ridge_stepwise <- MSPE_ridge_lasso <- vector(length = k)

# 10-Fold Cross-Validation for Ridge Regression
for (f in 1:k) {
  train <- (id != f)  # Training set
  valid <- (id == f)  # Validation set (within data_clean1)

  # Stepwise Ridge Model
  X_train_stepwise <- X_clean1_stepwise[train, ]
  X_valid_stepwise <- X_clean1_stepwise[valid, ]
  y_train_stepwise <- y_clean1[train]
  y_valid_stepwise <- y_clean1[valid]

  ridge_cv_stepwise <- cv.glmnet(X_train_stepwise, y_train_stepwise, alpha = 0, lambda = exp(seq(-20, 20, length = 200)))
  ridge_model_stepwise <- glmnet(X_train_stepwise, y_train_stepwise, lambda = ridge_cv_stepwise$lambda.min, alpha = 0)

  # Predict on `data`
  predictions_stepwise <- predict(ridge_model_stepwise, newx = X_test_stepwise)
  MSPE_ridge_stepwise[f] <- mean((predictions_stepwise - y_test)^2)

  # Lasso Ridge Model
  X_train_lasso <- X_clean1_lasso[train, ]
  X_valid_lasso <- X_clean1_lasso[valid, ]
  y_train_lasso <- y_clean1[train]
  y_valid_lasso <- y_clean1[valid]

  ridge_cv_lasso <- cv.glmnet(X_train_lasso, y_train_lasso, alpha = 0, lambda = exp(seq(-20, 20, length = 200)))
  ridge_model_lasso <- glmnet(X_train_lasso, y_train_lasso, lambda = ridge_cv_lasso$lambda.min, alpha = 0)

  # Predict on `data`
  predictions_lasso <- predict(ridge_model_lasso, newx = X_test_lasso)
  MSPE_ridge_lasso[f] <- mean((predictions_lasso - y_test)^2)
}

# Calculate the average MSPE
avg_MSPE_ridge_stepwise <- mean(MSPE_ridge_stepwise)
avg_MSPE_ridge_lasso <- mean(MSPE_ridge_lasso)

# Print results
cat("\nAverage MSPE for Ridge models trained on data_clean1 (with 10-fold CV) and tested on data:\n")
cat("Ridge with Stepwise features:", avg_MSPE_ridge_stepwise, "\n")
cat("Ridge with Lasso features:", avg_MSPE_ridge_lasso, "\n")
```

This code chunk is to save this model to the prediction data set. Sorry for spoilers but stepwise features with ridge regression was my best model. YOu can move past this!

```{r}
# Step 1: Read the data for prediction
data_te <- read.table("C:/Users/ashvi/OneDrive/Desktop/ML Business Forecasting/data_for_prediction.txt", header = TRUE, sep = "\t", dec = ".")

# Remove the 'tw' column from the data (it’s not part of the features for prediction)
data_te <- data_te[, setdiff(names(data_te), "tw")]

# Step 2: Prepare the test set matrix using selected features (including interaction terms)
# Use the same formula with interaction terms that was used during model training
formula_stepwise <- as.formula(paste("~", paste(selected_features_stepwise, collapse = " + ")))

# Create the model matrix for test data with interaction terms
X_test_te <- model.matrix(formula_stepwise, data = data_te)[, -1]  # Exclude intercept column

# Step 3: Check the number of features in X_test_te to match with the model's training data
cat("Number of features in test data:", ncol(X_test_te), "\n")

# Ensure the number of features matches the model's expectations (27 features in this case)
if (ncol(X_test_te) == 27) {
  # Step 4: Make predictions using the trained stepwise ridge model
  my_predictions <- predict(ridge_model_stepwise, newx = X_test_te)
  
  # Step 5: Add the predictions to the dataset as a new column
  data_te$predicted_tw <- my_predictions
  
  # Step 6: Save the updated data with predictions back to a new text file
  write.table(data_te, 
            "C:/Users/ashvi/OneDrive/Desktop/ML Business Forecasting/Ashvin_Predictions.txt", 
            sep = "\t", 
            row.names = FALSE, 
            col.names = TRUE)
  
  # Optional: Check the first few rows to confirm the added predictions
  head(data_te)

}
```

This is doing it with lasso regression, doesn't hurt to check. Ridge still performed best with the features selected with Step wise.
```{r}
# Function to calculate MSPE (Mean Squared Prediction Error)
calculate_mspe <- function(model, X_test, y_test) {
  predictions <- predict(model, newx = X_test)
  return(mean((predictions - y_test)^2))
}

# Set the number of folds for cross-validation (10-fold)
k <- 10
n <- nrow(data_clean1)  
set.seed(67134)
id <- sample(rep(1:k, length = n))

# Prepare Stepwise-selected features in `data_clean1`
X_clean1_stepwise <- model.matrix(as.formula(paste("tw ~", paste(selected_features_stepwise, collapse = " + "))), data = data_clean1)[, -1]
# Prepare Lasso-selected features in `data_clean1`
X_clean1_lasso <- model.matrix(as.formula(paste("tw ~", paste(selected_features_lasso, collapse = " + "))), data = data_clean1)[, -1]
y_clean1 <- data_clean1$tw

# Prepare test set from `data`
X_test_stepwise <- model.matrix(as.formula(paste("tw ~", paste(selected_features_stepwise, collapse = " + "))), data = data)[, -1]
X_test_lasso <- model.matrix(as.formula(paste("tw ~", paste(selected_features_lasso, collapse = " + "))), data = data)[, -1]
y_test <- data$tw

# Initialize vectors to store MSPE for each fold
MSPE_ridge_stepwise <- MSPE_ridge_lasso <- vector(length = k)

# 10-Fold Cross-Validation for Ridge Regression
for (f in 1:k) {
  train <- (id != f)  # Training set
  valid <- (id == f)  # Validation set (within data_clean1)

  # Stepwise Ridge Model
  X_train_stepwise <- X_clean1_stepwise[train, ]
  X_valid_stepwise <- X_clean1_stepwise[valid, ]
  y_train_stepwise <- y_clean1[train]
  y_valid_stepwise <- y_clean1[valid]

  ridge_cv_stepwise <- cv.glmnet(X_train_stepwise, y_train_stepwise, alpha = 1, lambda = exp(seq(-20, 20, length = 200)))
  ridge_model_stepwise <- glmnet(X_train_stepwise, y_train_stepwise, lambda = ridge_cv_stepwise$lambda.min, alpha = 1)

  # Predict on `data`
  predictions_stepwise <- predict(ridge_model_stepwise, newx = X_test_stepwise)
  MSPE_ridge_stepwise[f] <- mean((predictions_stepwise - y_test)^2)

  # Lasso Ridge Model
  X_train_lasso <- X_clean1_lasso[train, ]
  X_valid_lasso <- X_clean1_lasso[valid, ]
  y_train_lasso <- y_clean1[train]
  y_valid_lasso <- y_clean1[valid]

  ridge_cv_lasso <- cv.glmnet(X_train_lasso, y_train_lasso, alpha = 1, lambda = exp(seq(-20, 20, length = 200)))
  ridge_model_lasso <- glmnet(X_train_lasso, y_train_lasso, lambda = ridge_cv_lasso$lambda.min, alpha = 1)

  # Predict on `data`
  predictions_lasso <- predict(ridge_model_lasso, newx = X_test_lasso)
  MSPE_ridge_lasso[f] <- mean((predictions_lasso - y_test)^2)
}

# Calculate the average MSPE
avg_MSPE_ridge_stepwise <- mean(MSPE_ridge_stepwise)
avg_MSPE_ridge_lasso <- mean(MSPE_ridge_lasso)

# Print results
cat("\nAverage MSPE for Ridge models trained on data_clean1 (with 10-fold CV) and tested on data:\n")
cat("Ridge with Stepwise features:", avg_MSPE_ridge_stepwise, "\n")
cat("Ridge with Lasso features:", avg_MSPE_ridge_lasso, "\n")
```
##Flexible Linear Models
Now we will move onto flexible linear models. We will create a simple gam model to visualize the flexible relationship between continous variables
```{r}
# Fit the GAM model to generate the smooth terms
gam_model <- gam(tw ~ s(fsize) + s(inc) + s(educ) + s(age) + ira + nifa + male + e401 + twoearn + hmort + marr + hequity, data = data_clean1)

# Add the smooth terms (splines) to the dataset
data_clean1$s_fsize <- predict(gam_model, newdata = data_clean1, type = "terms")[, "s(fsize)"]
data_clean1$s_inc <- predict(gam_model, newdata = data_clean1, type = "terms")[, "s(inc)"]
data_clean1$s_educ <- predict(gam_model, newdata = data_clean1, type = "terms")[, "s(educ)"]
data_clean1$s_age <- predict(gam_model, newdata = data_clean1, type = "terms")[, "s(age)"]
```
We will put the transformed spline features into a dataset, and like before generate every possible 1:1 interaction term.


```{r}

# Get the original feature names (excluding 'tw')
original_features <- c("fsize", "inc", "educ", "age", "ira", "nifa", "male", "e401", "twoearn", "hmort")

# Generate all pairwise interaction terms between the original and smoothed features
interaction_terms <- combn(c(original_features, "s_fsize", "s_inc", "s_educ", "s_age"), 2, 
                           FUN = function(x) paste(x[1], x[2], sep=":"), simplify = TRUE)

# Create the formula for the linear model including both original features and interaction terms
formula_terms <- paste("tw ~", paste(c(original_features, "s_fsize", "s_inc", "s_educ", "s_age", interaction_terms), collapse=" + "))
print(formula_terms)

```

Now we will run feature selection method, stepwise both way ways to select features
```{r}
# Create the formula for the linear model including both original features and interaction terms
formula_terms <- paste("tw ~", paste(c(original_features, "s_fsize", "s_inc", "s_educ", "s_age", interaction_terms), collapse=" + "))
print(formula_terms)

#it full model with interaction terms on data_clean1
full_model <- lm(as.formula(formula_terms), data = data_clean1)

# Null model (intercept only)
null_model <- lm(tw ~ 1, data = data_clean1)

# Perform stepwise selection using both forward and backward directions based on AIC
stepwise_model <- stepAIC(null_model, scope = list(lower = null_model, upper = full_model), direction = "both", trace = FALSE)

# View selected features from the stepwise model
cat("Selected features in Stepwise Model:\n")
selected_features_stepwise <- coef(stepwise_model)[coef(stepwise_model) != 0]
selected_features_stepwise <- names(selected_features_stepwise)[selected_features_stepwise != "(Intercept)"]
print(selected_features_stepwise)
```
We will do it again with Lasso
```{r}
# Create the model matrix (predictors) for interaction terms
X_interactions <- model.matrix(as.formula(formula_terms), data = data_clean1)[,-1]  # Remove intercept column
y <- data_clean1$tw  # Target variable

set.seed(1234)
# Perform Lasso regression with cross-validation to select the optimal lambda
lasso_cv <- cv.glmnet(X_interactions, y, alpha = 1)  # alpha=1 for Lasso regression
cat("Best lambda from Lasso CV:", lasso_cv$lambda.min, "\n")

# Fit Lasso model using the best lambda
lasso_model <- glmnet(X_interactions, y, lambda = lasso_cv$lambda.min, alpha = 1)

# View the coefficients of the Lasso model
cat("\nLasso Regression Coefficients:\n")
coef(lasso_model)
```
Cross validation of ridge regression using lasso selected features and stepwise
```{r}
set.seed(67134)  # Set the seed for reproducibility

# Number of folds for 10-fold cross-validation
k <- 10
n <- nrow(data_clean1)  
id <- sample(rep(1:k, length = n))  # 10-fold cross-validation IDs

# For Lasso-selected features
X_lasso <- model.matrix(as.formula(paste("tw ~", paste(selected_features_lasso, collapse = " + "))), data = data_clean1)[,-1]
y_lasso <- data_clean1$tw  # Target variable for Lasso model

# Stepwise model fitting (you've already done this with your code)
selected_features_stepwise <- names(coef(stepwise_model)[coef(stepwise_model) != 0])
selected_features_stepwise <- selected_features_stepwise[selected_features_stepwise != "(Intercept)"]  # Remove intercept

# Step 2: Create the model matrix for Stepwise-selected features
X_stepwise <- model.matrix(as.formula(paste("tw ~", paste(selected_features_stepwise, collapse = " + "))), data = data_clean1)[,-1]  # Remove intercept
y_stepwise <- data_clean1$tw  # Target variable for Stepwise model

## Step 3: Now, perform Ridge regression for Lasso and Stepwise-selected features using 10-fold cross-validation
# Initialize vectors to store MSPE for both models
MSPE_ridge_lasso <- MSPE_ridge_stepwise <- vector(length = k)

# Cross-validation loop for Ridge Regression using 10-fold cross-validation
for (f in 1:k) {
  train <- (id != f)  # Training set
  valid <- (id == f)  # Validation set

  # Ridge model using Lasso-selected features
  ridge_cv_lasso <- cv.glmnet(X_lasso[train, ], y_lasso[train], alpha = 0)  # alpha=0 for Ridge regression
  ridge_model_lasso <- glmnet(X_lasso[train, ], y_lasso[train], lambda = ridge_cv_lasso$lambda.min, alpha = 0)
  predictions_lasso <- predict(ridge_model_lasso, newx = X_lasso[valid, ])
  MSPE_ridge_lasso[f] <- mean((predictions_lasso - y_lasso[valid])^2)

  # Ridge model using Stepwise-selected features
  ridge_cv_stepwise <- cv.glmnet(X_stepwise[train, ], y_stepwise[train], alpha = 0)  # alpha=0 for Ridge regression
  ridge_model_stepwise <- glmnet(X_stepwise[train, ], y_stepwise[train], lambda = ridge_cv_stepwise$lambda.min, alpha = 0)
  predictions_stepwise <- predict(ridge_model_stepwise, newx = X_stepwise[valid, ])
  MSPE_ridge_stepwise[f] <- mean((predictions_stepwise - y_stepwise[valid])^2)
}

# Step 4: Calculate the average MSPE for both models
avg_MSPE_ridge_lasso <- mean(MSPE_ridge_lasso)
avg_MSPE_ridge_stepwise <- mean(MSPE_ridge_stepwise)

cat("\nAverage MSPE for Ridge models trained on data_clean1 and tested on data:\n")
cat("Ridge with Lasso-selected features:", avg_MSPE_ridge_lasso, "\n")
cat("Ridge with Stepwise-selected features:", avg_MSPE_ridge_stepwise, "\n")

```
They performed much worse then the linear models. I will now try with polynomials and splines

```{r}
set.seed(123)  # For reproducibility

# Define the number of folds for cross-validation
k <- 10

# Split the data into k-folds (cross-validation)
n <- nrow(data_clean1)
folds <- sample(1:k, size = n, replace = TRUE)

# Set up a vector to store cross-validation errors for each polynomial degree
cv_errors_inc <- numeric(5)

# Perform cross-validation for polynomial degrees 1 to 5 (for inc)
for (degree in 1:5) {
  cv_error <- 0  # Initialize error for this degree
  
  for (fold in 1:k) {
    # Split the data into training and testing sets based on the fold
    train_data <- data_clean1[folds != fold, ]
    test_data <- data_clean1[folds == fold, ]
    
    # Create the formula with polynomial terms for inc (using I() for manual polynomials)
    formula <- as.formula(paste("tw ~", paste("I(inc^", 1:degree, ")", collapse = " + ")))
    
    # Fit the model using the polynomial terms
    model <- lm(formula, data = train_data)
    
    # Make predictions on the test set
    predictions <- predict(model, newdata = test_data)
    
    # Calculate the Mean Squared Error (MSE) for this fold
    mse <- mean((test_data$tw - predictions)^2)
    cv_error <- cv_error + mse
  }
  
  # Average the cross-validation error over all folds for this degree
  cv_errors_inc[degree] <- cv_error / k
}

# Print the cross-validation errors for each degree (for inc)
cv_errors_inc

# Choose the degree with the lowest cross-validation error for inc
best_degree_inc <- which.min(cv_errors_inc)
best_degree_inc







```

```{r}
# Load required package
library(splines)

# Cross-validation to choose the optimal degree for splines
set.seed(614)
y <- data_clean1$tw  # Dependent variable
n <- length(y)       # Number of observations
k <- 10              # Number of folds for cross-validation
id <- sample(rep(1:k, length = n))  # Cross-validation fold assignment

# Maximum number of df for splines
max_df <- 20
mspe_age <- rep(NA, max_df)
mspe_inc <- rep(NA, max_df)

# Cross-validation for age
for (deg in 1:max_df) {
  mspe <- vector(length = k)
  
  for (f in 1:k) {
    test <- (id == f)
    train <- (id != f)
    
    # Cubic spline fit for age with df = deg
    spline_age <- lm(tw ~ bs(age, df = deg), data=data_clean1[train,])
    
    # Make predictions
    pred <- predict(spline_age, newdata = data_clean1[test,])
    
    # Calculate MSPE
    mspe[f] <- mean((y[test] - pred)^2)
  }
  mspe_age[deg] <- mean(mspe)
}

# Cross-validation for income
for (deg in 1:max_df) {
  mspe <- vector(length = k)
  
  for (f in 1:k) {
    test <- (id == f)
    train <- (id != f)
    
    # Cubic spline fit for income with df = deg
    spline_inc <- lm(tw ~ bs(inc, df = deg), data=data_clean1[train,])
    
    # Make predictions
    pred <- predict(spline_inc, newdata = data_clean1[test,])
    
    # Calculate MSPE
    mspe[f] <- mean((y[test] - pred)^2)
  }
  mspe_inc[deg] <- mean(mspe)
}

# Find the optimal degree (minimum MSPE)
optimal_degree_age <- which.min(mspe_age)
optimal_degree_inc <- which.min(mspe_inc)

cat("Optimal degree for age:", optimal_degree_age, "\n")
cat("Optimal degree for income:", optimal_degree_inc, "\n")

# Fit final models with optimal degrees
spline_age_final <- lm(tw ~ bs(age, df = optimal_degree_age), data=data_clean1)
spline_inc_final <- lm(tw ~ bs(inc, df = optimal_degree_inc), data=data_clean1)

# Plotting the results
# For Age
plot(data_clean1$age, data_clean1$tw, xlab="Age", ylab="TW", main="Optimal Spline Regression for Age", pch=19, cex=0.5, col="red")
lines(sort(data_clean1$age), predict(spline_age_final)[order(data_clean1$age)], col="blue", lwd=2)

# For Income
plot(data_clean1$inc, data_clean1$tw, xlab="Income", ylab="TW", main="Optimal Spline Regression for Income", pch=19, cex=0.5, col="red")
lines(sort(data_clean1$inc), predict(spline_inc_final)[order(data_clean1$inc)], col="blue", lwd=2)

```


```{r}
# Create the polynomial and spline features for age and income
data_clean1$poly_age <- data_clean1$age  # 1st degree polynomial for age
data_clean1$spline_age <- predict(lm(tw ~ bs(age, df = 7), data = data_clean1), newdata = data_clean1)  # 7th degree spline for age

# Create 2nd-degree polynomial for income and 4th-degree spline for income
data_clean1$poly_inc <- predict(lm(tw ~ poly(inc, 2), data = data_clean1), newdata = data_clean1)  # 2nd degree polynomial for income
data_clean1$spline_inc <- predict(lm(tw ~ bs(inc, df = 4), data = data_clean1), newdata = data_clean1)  # 4th degree spline for income

# Prepare the data for Lasso: Select predictors and response variable
X_clean1 <- data_clean1[, c("age", "inc", "poly_age", "spline_age", "poly_inc", "spline_inc")]
y_clean1 <- data_clean1$tw

# Scaling the predictors (important for Lasso)
X_clean1_scaled <- scale(X_clean1)

# Cross-validation setup
set.seed(614)
n <- length(y_clean1)       # Number of observations
k <- 10                     # Number of folds for cross-validation
id <- sample(rep(1:k, length = n))  # Cross-validation fold assignment

# Initialize MSPE vector for each model
mspe_spline_age <- rep(NA, k)
mspe_spline_inc <- rep(NA, k)
mspe_poly_age <- rep(NA, k)
mspe_poly_inc <- rep(NA, k)

# Perform 10-fold cross-validation for each model
for (f in 1:k) {
  test <- (id == f)
  train <- (id != f)
  
  # Get training and test data for both sets
  X_train <- X_clean1_scaled[train, ]
  y_train <- y_clean1[train]
  X_test <- X_clean1_scaled[test, ]
  y_test <- y_clean1[test]
  
  # Model 1: Spline for Age (7 degrees)
  spline_age_model <- lm(tw ~ bs(age, df = 7), data = data_clean1[train,])
  pred_spline_age <- predict(spline_age_model, newdata = data_clean1[test,])
  mspe_spline_age[f] <- mean((y_test - pred_spline_age)^2)
  
  # Model 2: Spline for Income (4 degrees)
  spline_inc_model <- lm(tw ~ bs(inc, df = 4), data = data_clean1[train,])
  pred_spline_inc <- predict(spline_inc_model, newdata = data_clean1[test,])
  mspe_spline_inc[f] <- mean((y_test - pred_spline_inc)^2)
  
  # Model 3: Linear Age (No transformation, just the raw data)
  poly_age_model <- lm(tw ~ age, data = data_clean1[train,])
  pred_poly_age <- predict(poly_age_model, newdata = data_clean1[test,])
  mspe_poly_age[f] <- mean((y_test - pred_poly_age)^2)
  
  # Model 4: Polynomial for Income (2nd degree)
  poly_inc_model <- lm(tw ~ poly(inc, 2), data = data_clean1[train,])
  pred_poly_inc <- predict(poly_inc_model, newdata = data_clean1[test,])
  mspe_poly_inc[f] <- mean((y_test - pred_poly_inc)^2)
}

# Calculate the average MSPE for each model
mean_mspe_spline_age <- mean(mspe_spline_age)
mean_mspe_spline_inc <- mean(mspe_spline_inc)
mean_mspe_poly_age <- mean(mspe_poly_age)
mean_mspe_poly_inc <- mean(mspe_poly_inc)

# Print the results
cat("Mean MSPE for Spline Age (7th degree):", mean_mspe_spline_age, "\n")
cat("Mean MSPE for Spline Income (4th degree):", mean_mspe_spline_inc, "\n")
cat("Mean MSPE for Linear Age:", mean_mspe_poly_age, "\n")
cat("Mean MSPE for Polynomial Income (2nd degree):", mean_mspe_poly_inc, "\n")

```

I ran lasso to select which features to run in my flexible linear model. None of the poly features were chosen. I would argue with the information from all the cross validations & average MSPE, that in the goal for prediction, none of the transformations I have done would improve the MSPE from my current best model. 
```{r}
# Step 1: Create the polynomial feature for income (2nd-degree polynomial)
data_clean1$poly_inc <- predict(lm(tw ~ poly(inc, 4), data = data_clean1), newdata = data_clean1)  # 2nd-degree polynomial for income

# Step 2: Prepare the feature matrix (include original features + polynomial feature)
# Assuming 'inc' is in the dataset, and 'tw' is the target variable
X_full <- as.matrix(cbind(1, data_clean1[, -which(names(data_clean1) == "tw")], data_clean1$poly_inc))  # Add intercept and poly_inc

y <- data_clean1$tw  # Target variable

# Step 3: Run Lasso regression on the full dataset (original features + polynomial income feature)
lasso_cv <- cv.glmnet(X_full, y, alpha = 1, lambda = exp(seq(-20, 20, length = 200)))

# Step 4: Get the coefficients from the Lasso model
lasso_coefficients <- coef(lasso_cv, s = "lambda.min")

# Step 5: Print the coefficients to see which features were selected
print(lasso_coefficients)

# Step 6: Extract the non-zero features (coefficients)
selected_features <- which(lasso_coefficients != 0)[-1]  # Exclude the intercept
print(selected_features)  # Check the selected features

# Check the selected features to ensure there are valid indices
if (length(selected_features) == 0) {
  stop("No features were selected by Lasso.")
} else {
  cat("Selected Features: ", selected_features, "\n")
}
```
